---
title: "Quick Start with dssrip2"
author: "Michael Koohafkan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
options(max.print = 10)
```

The R package `dssrip2` provides a simple interface to read and write
HEC-DSS files.

## Connecting to HEC-DSSVue

If you have HEC-DSSVue installed on your machine, call `dss_connect()`
and supply the path to the install folder:

```{r eval = FALSE}
library(dssrip2)

dss_connect("C:/Program Files/HEC/HEC-DSSVue")
```

Alternatively, you can download the HEC Monolith Jars (and associated
dependencies) and connect to those instead:

```{r eval = FALSE}
dss_install_monolith()
dss_connect(monolith = TRUE)
```

By default, HEC Monolith installs to APPDATA. For simplified startup,
you can add the following options to `.Rprofile`:

```{r eval = FALSE}
# for HEC-DSSVue
options(dss.home = "C:/Program Files/HEC/HEC-DSSVue")

# alternatively, for HEC Monolith in the default direcory
options(dss.monolith = TRUE)
```

Review the `dss_connect()` and `dss_install_monolith()` help files for
more information on configuration.


## Reading DSS data

Start by opening a DSS File:

```{r}
f = dss_open(system.file("extdata/example.dss", package = "dssrip2"))
```

This file contains five datasets:

1. A daily flow timeseries
2. A 15-minute stage timeseries
3. An annual peak flow irregular timeseries
4. An annual peak stage irregular timeseries
5. A flow-elevation paired datasets

The DSS catalog can be read into R using `dss_catalog()`:

```{r}
dss_catalog(f, condensed = TRUE)
```

`dss_catalog()` supports regex pattern matching to subset the path list:

```{r}
dss_catalog(f, "/.*/.*/FLOW.*/.*/.*/.*/")
```

`dssrip2` also provides a variety of functions are provided to assist
with manipulating DSS paths:

```{r}
path = "/BRANDYWINE CREEK/WILMINGTON, DE/FLOW/01Oct1946 - 30Sep2012/1DAY/USGS/"

# split path into parts
dss_parts_split(path)

# replace path part
dss_parts_replace(path, list(B = "WILMINGTON, DELAWARE"))

# combine parts into path
dss_parts_combine(list(A = "BASIN", B = "LOCATION", C = "PARAMETER",
  D = "", E = "1DAY", F = "EXAMPLE"))
```

`dssrip2` provides the function `dss_read()` to read DSS datasets into
R. The output format is automatically deduced from the DSS data type:

- DSS regular and irregular time series are converted into two-column data frames.
- DSS paired datasets are converted into two-column (or more) data frames.
- DSS grid data is converted into SpatRasters
  (see [`package:terra`](https://cran.r-project.org/package=terra)
  for more information).

```{r}
# read in a regular timeseries
dss_read(f, "/BRANDYWINE CREEK/WILMINGTON, DE/FLOW//1DAY/USGS/")

# read in a single chunk of a regular timeseries
dss_read(f, "/BRANDYWINE CREEK/WILMINGTON, DE/FLOW/01JAN1948/1DAY/USGS/",
  full = FALSE)

# read in irregular timeseries
dss_read(f, "/BRANDYWINE CREEK/WILMINGTON, DE/FLOW-ANNUAL PEAK//IR-CENTURY/USGS/")

# read in a paired dataset
dss_read(f, "/BRANDYWINE CREEK/WILMINGTON, DE/FLOW-STAGE///GENERATED DATA PAIRS/")
```


## Writing DSS data

`dssrip2` Also provides write support for time series and paired data
containers. As with reading data, `dssrip2` determines the appropriate
DSS dataset type from the structure of the supplied input.

```{r}
outfile = tempfile(fileext = ".dss")

f = dss_create(outfile)

# create a regular timeseries from the LakeHuron dataset
data(LakeHuron)
lake_huron = data.frame(date = as.Date(sprintf("%d-10-01",
  time(LakeHuron))), elev = as.vector(LakeHuron))
lake_huron_path = "/GREAT LAKES/LAKE HURON/ELEV//1YEAR/R DATASET/"
lake_huron_meta = list(units = "feet", type = "INST")

dss_write(lake_huron, f, lake_huron_path, lake_huron_meta)

# create an irregular timeseries from the airquality dataset
data(airquality)
aq = do.call(rbind, lapply(split(airquality, airquality$Month),
  function(x)
    x[which.max(x$Temp), ]))
temps = data.frame(date = as.Date(sprintf("1973-%d-%d",
  aq$Month, aq$Day)), temperature = aq$Temp)
temps_path = "/NEW YORK/NEW YORK CITY/TEMPERATURE//IR-YEAR/R DATASET/"
temps_meta = list(units = "degrees F", type = "INST")

dss_write(temps, f, temps_path, temps_meta)

# create paired dataset from the Loblolly dataset
data(Loblolly)
growth = reshape(Loblolly, direction = "wide", idvar = "age", timevar = "Seed")
names(growth) = gsub("height.", "seed_", names(growth))
growth_path = "//LOBLOLLY PINES/AGE-HEIGHT///R DATASET/"
growth_meta = list(xunits = "years", yunits = "feet",
  xtype = "UNT", ytype = "UNT", labels = names(growth)[2:ncol(growth)])

dss_write(growth, f, growth_path, growth_meta)
```

That's it!

