---
title: "Quick Start with dssrip2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
options(max.print = 10)
```

The R package `dssrip2` provides a simple interface to read and write
HEC-DSS files.

## Settings Java Options

The `dssrip2` uses the `rJava` package, which provides a Java Virtual
Machine (JVM) shared by all R packages in a session. The default JVM
parameters are very modest and usually insufficient for working with
larger DSS files. Before launching `dssrip2` (or any other R package
that uses `rJava`, as the JVM is launched by the first package that
requires it) you should at least set the JVM total memory
parameter. For example, set the JVM total memory to 8000 MB:

```{r eval = FALSE}
options(java.parameters = "-Xmx8000M")
```

For simplified startup, set the options in your `.Rprofile`.
For more information on setting Java parameters, see the 
*Java Parameters* vignette.

## Connecting to DSS

### Using HEC Monolith

The recommended method of connecting to DSS is to download the HEC
Monolith jar files, which provides standalone access to HEC DSS
functionality without requiring the installation of the HEC-DSSVue
application.

```{r eval = FALSE}
dss_install_monolith()
dss_connect(monolith = TRUE)
```

By default, HEC Monolith installs to APPDATA. For simplified startup,
you can add the following options to `.Rprofile`:

```{r eval = FALSE}
options(dss.monolith = TRUE)
```

### Using an existing HEC-DSSVue install

Alternatively, if you have HEC-DSSVue installed on your machine you
can call `dss_connect()` and supply the path to the DSSVue install
folder:

```{r eval = FALSE}
library(dssrip2)

dss_connect("C:/Program Files/HEC/HEC-DSSVue")
```

For simplified startup, you can add the following options
to `.Rprofile`:

```{r eval = FALSE}
options(dss.home = "C:/Program Files/HEC/HEC-DSSVue",
  dss.monolith = FALSE)
```


## Reading DSS data

`dssrip2` provides an example dataset:

```{r}
exfile = system.file("extdata/example.dss", package = "dssrip2")
```

This file contains five datasets:

1. A daily flow timeseries
2. A 15-minute stage timeseries
3. An annual peak flow irregular timeseries
4. An annual peak stage irregular timeseries
5. A flow-elevation paired datasets

The DSS catalog can be read into R using `dss_catalog()`:

```{r}
dss_catalog(exfile, condensed = TRUE)
```

`dss_catalog()` supports regex pattern matching to subset the path list:

```{r}
dss_catalog(exfile, "/.*/.*/FLOW.*/.*/.*/.*/")
```

`dssrip2` also provides a variety of functions are provided to assist
with manipulating DSS paths:

```{r}
path = "/BRANDYWINE CREEK/WILMINGTON, DE/FLOW/01Oct1946 - 30Sep2012/1DAY/USGS/"

# split path into parts
dss_parts_split(path)

# replace path part
dss_parts_replace(path, list(B = "WILMINGTON, DELAWARE"))

# combine parts into path
dss_parts_combine(list(A = "BASIN", B = "LOCATION", C = "PARAMETER",
  D = "", E = "1DAY", F = "EXAMPLE"))
```

`dssrip2` provides the function `dss_read()` to read DSS datasets into
R. The output format is automatically deduced from the DSS data type:

- DSS regular and irregular time series are converted into two-column data frames.
- DSS paired datasets are converted into two-column (or more) data frames.
- DSS grid data is converted into SpatRasters
  (see [`package:terra`](https://cran.r-project.org/package=terra)
  for more information).

```{r}
# read in a regular timeseries
dss_read(exfile, "/BRANDYWINE CREEK/WILMINGTON, DE/FLOW//1DAY/USGS/")

# read in a single chunk of a regular timeseries
dss_read(exfile, "/BRANDYWINE CREEK/WILMINGTON, DE/FLOW/01JAN1948/1DAY/USGS/",
  full = FALSE)

# read in irregular timeseries
dss_read(exfile, "/BRANDYWINE CREEK/WILMINGTON, DE/FLOW-ANNUAL PEAK//IR-CENTURY/USGS/")

# read in a paired dataset
dss_read(exfile, "/BRANDYWINE CREEK/WILMINGTON, DE/FLOW-STAGE///GENERATED DATA PAIRS/")
```


## Writing DSS data

`dssrip2` Also provides write support for time series and paired data
containers. As with reading data, `dssrip2` determines the appropriate
DSS dataset type from the structure of the supplied input.

```{r}
outfile = tempfile(fileext = ".dss")

dss_create(outfile)

# create a regular timeseries from the LakeHuron dataset
data(LakeHuron)
lake_huron = data.frame(date = as.Date(sprintf("%d-10-01",
  time(LakeHuron))), elev = as.vector(LakeHuron))
lake_huron_path = "/GREAT LAKES/LAKE HURON/ELEV//1YEAR/R DATASET/"

# additional attributes are required to write DSS data
lake_huron = dss_add_attributes(lake_huron, list(units = "feet",
  type = "INST"))

dss_write(lake_huron, outfile, lake_huron_path)

# create an irregular timeseries from the airquality dataset
data(airquality)
aq = do.call(rbind, lapply(split(airquality, airquality$Month),
  function(x)
    x[which.max(x$Temp), ]))
temps = data.frame(date = as.Date(sprintf("1973-%d-%d",
  aq$Month, aq$Day)), temperature = aq$Temp)
temps_path = "/NEW YORK/NEW YORK CITY/TEMPERATURE//IR-YEAR/R DATASET/"

# additional attributes
temps = dss_add_attributes(temps, list(units = "degrees F",
  type = "INST"))

dss_write(temps, outfile, temps_path)

# create paired dataset from the Loblolly dataset
data(Loblolly)
growth = reshape(Loblolly, direction = "wide", idvar = "age", timevar = "Seed")
names(growth) = gsub("height.", "seed_", names(growth))
growth_path = "//LOBLOLLY PINES/AGE-HEIGHT///R DATASET/"

# additional attributes
growth = dss_add_attributes(growth, list(xunits = "years",
  yunits = "feet", xtype = "UNT", ytype = "UNT"))

dss_write(growth, outfile, growth_path)
```

It is good practice to close a DSS file once you are finished working
with it. For the most part, `dssrip2` will call the file's `$done()`
method at the completion of every operation to implicitly close the
file, but you can also explicitly close a file:

```{r}
dss_close(outfile)

# or close all open/implicitly closed files
dss_close_all()
```

That's it!
